SCATER CASE STUDY

### 2016-08-14 scater_case_study/case_study.Rmd
scater paper ready for submission

---
title: "SCATER CASE STUDY"
author: Davis McCarthy & Quin Wills
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        theme: united
        highlight: tango
        number_sections: false
        code_folding: hide
---

@@

# setwd( "~/Dropbox/scater_case_study/" )
# setwd( "~/Google\ Drive/Manuscripts/scater_paper_nov-2015/scater_case_study/" )
library( "scater" )
library( "EBImage" )
library( "scran" )
library( "RUVSeq" )
```

1. **OVERVIEW:** Brief introduction to scater and the case study data.
2. **GETTING STARTED:** Import data/meta-data,  and run automated QC.
3. **QC THE CELLS:** Exploratory plots and quality control of the cells.  
4. **QC THE GENES:** Exploratory plots and quality control of the genes.  
5. **NORMALISE THE DATA**: Unwanted technical effects in the data, and corrective normalisation.  
6. **DOWNSTREAM ANALYSIS:** An example of further analysis.
7. **TECHNICAL STUFF:** Information for bioinformaticians, such as the R session used to run this case study.

@@

### 1. OVERVIEW

@@

### 2. GETTING STARTED

@@

In keeping with best practice in object-oriented bioinformatics, `scater` not only allows you to store your data in a standardised format, but also to associate it with experimental and bioinformatic meta-data. Experimental meta-data are any information associated with individual cells (also called 'samples'), from sequencing metrics to experimental batching and other experimental cell phenotypes. The bioinformatic meta-data are any information related to the genes (also called 'features'). These can be read in,  combined with the sequencing data,  and extracted as shown below. For this example we'll use the automated log-scaled counts per million generated by `scater`,  but see ```calculateTPM()``` as an example function for generating alternative RNA-seq read count formats.  
\

View the first few entris for the cell metadata:

@@

cells <- AnnotatedDataFrame( read.csv( "cells.csv", row.names = 1 ) ) # row names match column names in reads
genes <- AnnotatedDataFrame( read.csv( "genes.csv", row.names = 1 ) ) # row names match row names in reads
scData <- newSCESet( countData = reads, phenoData = cells, featureData = genes, lowerDetectionLimit = 3 )

pData( scData )[1:3, 1:2] # view some sample meta-data
```

@@

### 3. QC THE CELLS

@@

```
\  
Another option available in `scater` is to conduct PCA on a set of QC metrics.
The advantage of doing this is that the QC metrics focus on technical aspects of
the libraries that are likely to distinguish problematics cells. Automatic 
outlier detection on PCA plots using QC metrics is available to help identify 
potentially problematic cells.

@@

plotReducedDim(scData, colour_by = "outlier", shape_by = "use", ncomponents = 3)

plotReducedDim(scData[, grep("cell", scData$sample_type)], 
               colour_by = "outlier", shape_by = "use", ncomponents = 3)

@@

With `scater`, any specific set of features based on prior knowledge can be used for PCA, t-SNE or diffusion maps. A feature set to use can be defined by supplying the `feature_set` argument to `plotPCA`, `plotTSNE` or `plotDiffusionMap`. This allows, for example, using only housekeeping features or control features or cell cycle genes to produce reduced-dimension plots. The plots below use only the spike-in genes defined as such earlier.

@@

```{r pca-diffmap-ccn-genes, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center"}
cell_cycle_genes <- !is.na(fData(scData)$cell_cycle_phase)
sum(cell_cycle_genes)
most_exprs_ccycle <- which(cell_cycle_genes & 
                               fData(scData)$total_feature_counts > 100000)
knitr::kable(fData(scData)[most_exprs_ccycle,] )
p1 <- plotPCA(scData, feature_set = cell_cycle_genes, 
               colour_by = featureNames(scData)[most_exprs_ccycle[8]], 
               shape_by = "sample_type", 
               size_by = "use" ) + ggtitle("PCA")
p2 <- plotDiffusionMap( scData, feature_set = cell_cycle_genes, 
                        colour_by = featureNames(scData)[most_exprs_ccycle[8]], 
                        shape_by = "sample_type", 
                        size_by = "use" ) + ggtitle("Diffusion map")
multiplot(p1, p2, cols = 2)

@@

One element of QC that can prove challenging to judge from the above criteria is if more than one cell ( or cell debris from other cells ) are being sequenced. With most cell processing platforms a small percentage of cells won't be truly single. One solution is imaging individual cells,  which also allows for the incorporation of fluorescence phenotypes such as viability and nuclear content ( for cell cycle analysis ). Using the [EBImage]( http://bioconductor.org/packages/release/bioc/html/EBImage.html ) package,  an example image of a stained cell is processed and displayed below,  with results incorporated into the sample meta-data.  
\

@@

### 4. QC THE GENES

@@

It can also be useful to look at total expression levels acrsoo gene biotypes. 
The plot below shows the distributions of log-10 scale total gene counts for 
genes of different annotated biotypes. The biotypes are ordered by median total
counts, and the jittered points beneath the violin plots indiciate the number of 
genes in the biotype. We want to see many protein-coding genes with reasonably 
high expression. This is exactly what we see here, with protein-coding genes 
appearing as the biotype with second-highest median expression behind the very 
small (but highly expressed) category of Mt_rRNA genes.
\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=8, fig.height=6, fig.align="center"}
oo <- order(unlist(dplyr::summarise(dplyr::group_by(fData(scData), gene_biotype),
                 median(log10_total_feature_counts))[,2]), decreasing = TRUE)
fData(scData)$biotype <- factor(fData(scData)$gene_biotype, 
                                levels = levels(fData(scData)$gene_biotype)[oo])

@@

```{r append-gene-symbols}
featureNames(scData) <- paste(fData(scData)$hgnc_symbol, featureNames(scData),
                              sep = "_")
```

```{r, echo=T, eval=T, message=F, warning=F, fig.width=6, fig.height=6, fig.align="center"}
plotQC( scData[fData( scData )$use, pData( scData )$use], 
        type = "highest-expression", 
        col_by_variable = "sample_type" )
```

@@

### 5. NORMALISE THE DATA

@@

Experimental design is a critical, but neglected, aspect of \gls{scrnaseq} studies. To the best of my knowledge, methods like those described in this section for exploring experimental and QC variables and the experimental design, do not feature in any scRNA-seq software packages apart from `scater`. There are a very large number of potential confounders, artifacts and biases in \gls{scrnaseq} studies. Exploring the effects of such explanatory variables (both those recorded during the experiment and computed QC metrics) is crucial for appropriate modeling of the data. The \scater\ package provides a set of methods specifically for quality control of experimental and explanatory variables, which will be demonstrated briefly here.

@@

In `scater` it is easy to perform size-factor normalisation using the "TMM" approach (the default in the [edgeR]() package), the "RLE" approach (default in the [DESeq]() package), an "upper-quartile approach" (as proposed by Bullard et al (2010)) or simple scaling by total counts. Size-factor normalisation can be carried out on a subsetted SCESet object, normalising either to ERCC spike-in genes or to endogenous genes if desired. Under certain circumstances either may be appropriate. Careful thought should be given to applying scale-factor normalisation as underlying assumptions from methods developed for bulk RNA-seq may not be appropriate for all single-cell datasets.

However, [Lun et al (2016)](http://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0947-7) recently published a size-factor normalisation method specifically designed
for scRNA-seq data. This method performs much better on single-cell data and its
implementation in the Bioconductor package `scran` allows seamless integration 
into the `scater` workflow. (The `scran` package itself depends on `scater`).

We recommend using the `scran` size-factor normalisation approach and 
demonstrate its use here.

@@

scDataQC_spikenorm <- computeSumFactors(scDataQC, sizes = 10, 
                              get.spikes = TRUE)
summary(scDataQC_spikenorm$size_factor)
scDataQC_spikenorm <- normalise(scDataQC_spikenorm)

@@

We would prefer such technical effects to be removed by normalisation, so below we demonstrate a further alternative approach. We see how "customised" normalisation approaches can be easily incorporated into the `scater` workflow.

In this example we normalise and standardise counts conditioned on expression level, showing the PCA for the corrected data. One advantage of this approach is that a biologically 'noisy' gene is naturally defined as one with greater dispersion than other genes at a similar expression level. In the normalised data these are genes with a mean absolute deviation greater than 1.

@@

@@

```{r regress-factors-norm,  echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=9}
design <- model.matrix(~scDataQC$log10_counts_endogenous_features +
                           scDataQC$pct_counts_top_100_features +
                           scDataQC$pct_counts_feature_controls +
                           scDataQC$c1_machine)
set_exprs(scDataQC, "norm_exprs_resid") <- norm_exprs(
    normaliseExprs(scDataQC, design = design,
                   method = "none", exprs_values = "exprs",
                   return_norm_as_exprs = FALSE) )

@@

Here, regressing out only the technical QC metrics does not remove the major 
effect of C1 machine, but regressing out cDNA concentration as well does (so 
the first principal component now separates the two patients). Thus, for 
removing known effects, the regression approach implemented in `scater` is 
effective, but more needs to be done to find _hidden_ effects such as unknown
batch effects, or similar.

@@

@@

scDataRUV <- filter(scData, use)
scDataRUV <- scDataRUV[fData(scData)$use | fData(scData)$is_feature_control,]
scIdx <- matrix(-1, ncol = max(table(scDataRUV$sample_type)), nrow = 2)
scIdx[1,] <- which(scDataRUV$sample_type == "cell from patient B")
tmp <- which(scDataRUV$sample_type == "cell from patient A")
scIdx[2, 1:length(tmp)] <- tmp
# cIdx <- (fData(scDataRUV)$is_feature_control & fData(scDataRUV)$n_cells_exprs >= 10) 
## Running RUVs
cIdx <- rownames(scDataRUV)
ruv_list <- list()
for (k in 1:9) {
    ruvs <- RUVs(counts(scDataRUV), cIdx, k = k,
                 scIdx = scIdx, isLog = FALSE)
    norm_counts(scDataRUV) <- ruvs$normalizedCounts
    exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                           prior.count = scDataRUV@logExprsOffset)
    plt_norm_ruv <- plotPCA(
        scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = TRUE) + 
        ggtitle(paste("PCA - RUVs normalisation: k =", k)) + 
        theme_bw(8)
    ruv_list[[k]] <- plt_norm_ruv
}

figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list[1:8], labels = letters[1:8], ncol = 2)
figplot_ruv
figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list, labels = letters[1:9])
cowplot::save_plot("../figures/supplement/ruvs_figure.png", figplot_ruv, 
                   ncol = 3, nrow = 3, base_height = 5, base_aspect_ratio = 0.9)
```

```{r ruv-seq-s-qc-plots, echo=TRUE, eval=TRUE, message=FALSE, results='hide'}
plotQC(scDataRUV, type = "explanatory-variables")
plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
```

```{r ruvseq-red-dim-k6}
ruvs <- RUVs(counts(scDataRUV), cIdx, k = 6, scIdx = scIdx, isLog = FALSE)
reducedDimension(scDataRUV) <- ruvs$W
plotReducedDim(scDataRUV, shape_by = "c1_machine", 
               colour_by = "sample_type", size_by = "total_features") + theme_bw()

Look at the performance of the RUVg method:

```{r ruv-seq-g, echo=TRUE, eval=TRUE, fig.height=14, fig.width=9}
# One approach to finding and removing hidden, unwanted sources of variability is
# the `RUVSeq` package.
## Running RUVs
cIdx <- fData(scDataRUV)$is_feature_control
ruvg_list <- list()
for (k in 1:9) {
    ruvs <- RUVg(counts(scDataRUV), cIdx, k = k, isLog = FALSE)
    norm_counts(scDataRUV) <- ruvs$normalizedCounts
    exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                           prior.count = scDataRUV@logExprsOffset)
    plt_norm_ruv <- plotPCA(
        scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = TRUE) + 
        ggtitle(paste("PCA - RUVg normalisation: k =", k)) + 
        theme_bw(8)
    ruvg_list[[k]] <- plt_norm_ruv
}

figplot_ruvg <- cowplot::plot_grid(plotlist = ruvg_list[1:8], labels = letters[1:8],
                                   ncol = 2)
figplot_ruvg
figplot_ruvg <- cowplot::plot_grid(plotlist = ruvg_list, labels = letters[1:9])
cowplot::save_plot("../figures/supplement/ruvg_figure.png", figplot_ruv, 
                   ncol = 3, nrow = 3, base_height = 5, base_aspect_ratio = 0.9)

```{r ruv-g-qc-plots, message=FALSE, results='hide'}
plotQC(scDataRUV, type = "explanatory-variables")
plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
scDataRUV$c1_machine <- as.factor(scDataRUV$c1_machine)
plotQC(scDataRUV, type = "find-pcs", variable = "c1_machine")

Unfortunately, the RUVg method does not do a good job of removing the C1 machine effect. The machine used is still correlated with the first PC even after removing 9 hidden factors of unwanted variation.

@@

### 7. DOWNSTREAM ANALYSIS

@@

In `scater` the `plotExpression` function enables the convenient visualisation of expression values for a set of features. Here, the expression values for the six most DE genes for expression frequency between patients are shown. The units for expression in the plot can be defined with the `exprs_values` argument (the expression values must exist with the provided name in the `assayData` slot of the SCESet object; if not the default `exprs` values will be used, with a warning). As with other plots in `scater` we can use phenotype data variables to define the colour and shapr of the points.

@@

### 8.  TECHNICAL STUFF

@@

```{r figures-for-paper, echo=TRUE, results='hide'}
### Figures for the paper
if ( requireNamespace("cowplot") ) {
    ## QC figure
    scData$well_type <- "single cell"
    scData$well_type[scData$sample_type == "blank (empty site) control"] <-
        "blank control"
    scData$well_type[scData$sample_type == "bulk control from patient A"] <-
        "bulk control"
    scData$well_type[scData$sample_type == "bulk control from patient B"] <-
        "bulk control"
    p1 <- plot( scData, colour_by = "well_type", exprs_values = "counts")
    p2 <- plotPCA( scData, shape_by = "sample_type", 
                   pca_data_input = "pdata", detect_outliers = TRUE,
                   selected_variables = c("cdna_recovered_in_ng_per_ul_no_miss",
                                          "pct_counts_top_100_features",
                                          "total_features", 
                                          "pct_counts_feature_controls",
                                          "n_detected_feature_controls", 
                                          "log10_counts_endogenous_features",
                                          "log10_counts_feature_controls"))
    p3 <- plotQC( scData[fData( scData )$use, pData( scData )$use], 
        type = "highest-expression", 
        col_by_variable = "sample_type" ) + coord_cartesian(xlim = c(0, 6))
    p4 <- plotQC(scData, type = "exprs") 
    p5 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "explanatory-variables", 
        variables = c("pct_counts_top_100_features", "total_features", 
                      "c1_machine",
                      "n_detected_feature_controls", 
                      "log10_counts_endogenous_features",
                      "sample_type") ) + 
        xlab(expression(paste(R ^ 2, " (% variance explained; log10-scale)"))) +
    theme(legend.position = "bottom", legend.text = element_text(size = 7))
    p6 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "c1_machine")
    figplot1 <- cowplot::plot_grid(p1, p2, p3, p4, p5, p6, 
                                  labels = letters, ncol = 3,
                                  label_size = 20)
    cowplot::save_plot("../figures/figure3.pdf", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3.png", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3.jpg", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4, dpi = 350)

    ## dimension reduction figure
    ww <- which(featureNames(scDataQC) %in% rownames(de)[c(1:2, 5)])
    # ww <- which(featureNames(scDataQC) %in% c("IGHM_ENSG00000211899", 
    #                                           "IGHV4-34_ENSG00000211956",
    #                                           "CD86_ENSG00000114013"))
    # ww <- which(featureNames(scDataQC) %in% c("IGHM", 
    #                                           "IGHV4-34",
    #                                           "CD86"))
    featureNames(scDataQC)[ww] <- as.character(fData(scDataQC)$hgnc_symbol[ww])
    feats_to_plot <- as.character(fData(scDataQC)$hgnc_symbol[ww])

>>>>

---
title: "SCATER CASE STUDY"
author: Davis McCarthy, Kieran Campbell, Aaron Lun & Quin Wills
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        theme: united
        highlight: tango
        number_sections: false
        code_folding: show
---

@@

# setwd( "~/Dropbox/scater_case_study/" )
# setwd( "~/Google\ Drive/Manuscripts/scater_paper_nov-2015/scater_case_study/" )
library( "scater" )
library( "EBImage" )
library( "scran" )
library( "RUVSeq" )
library( "BiocStyle" )
options(stringsAsFactors = FALSE)
```

1. **OVERVIEW:** Brief introduction to scater and the case study data.
2. **GETTING STARTED:** Import data/meta-data,  and run automated QC.
3. **QC THE CELLS:** Exploratory plots and quality control of the cells.  
4. **QC THE GENES:** Exploratory plots and quality control of the genes.  
5. **NORMALISE THE DATA**: Unwanted technical effects in the data, and corrective normalisation.  
6. **DOWNSTREAM ANALYSIS:** An example of further analysis.
7. **SESSION INFORMATION:** Information for bioinformaticians, such as the R session used to run this case study.

@@

## 1. OVERVIEW

@@

## 2. GETTING STARTED

@@

In keeping with best practice in object-oriented bioinformatics, `r Biocpkg("scater")` not only allows you to store your data in a standardised format, but also to associate it with experimental and bioinformatic meta-data. Experimental meta-data are any information associated with individual cells (also called 'samples'), from sequencing metrics to experimental batching and other experimental cell phenotypes. The bioinformatic meta-data are any information related to the genes (also called 'features'). These can be read in,  combined with the sequencing data,  and extracted as shown below. For this example we'll use the automated log-scaled counts per million generated by `r Biocpkg("scater")`,  but see ```calculateTPM()``` as an example function for generating alternative RNA-seq read count formats.  
\

View the first few entries for the cell metadata:

@@

cells <- AnnotatedDataFrame( read.csv( "cells.csv", row.names = 1 ) ) # row names match column names in reads
genes <- AnnotatedDataFrame( read.csv( "genes.csv", row.names = 1 ) ) # row names match row names in reads
scData <- newSCESet( countData = reads, phenoData = cells, featureData = genes, lowerDetectionLimit = 3 ) 
head(pData(scData))  # view some sample meta-data
```

@@

## 3. QC THE CELLS

@@

```
\  
Another option available in `r Biocpkg("scater")` is to conduct PCA on a set of QC metrics.
The advantage of doing this is that the QC metrics focus on technical aspects of
the libraries that are likely to distinguish problematics cells. Automatic 
outlier detection on PCA plots using QC metrics is available to help identify 
potentially problematic cells.

@@

@@

With `r Biocpkg("scater")`, any specific set of features based on prior knowledge can be used for PCA, t-SNE or diffusion maps. A feature set to use can be defined by supplying the `feature_set` argument to `plotPCA`, `plotTSNE` or `plotDiffusionMap`. This allows, for example, using only housekeeping features or control features or cell cycle genes to produce reduced-dimension plots. The plots below use only the spike-in genes defined as such earlier.

@@

```{r pca-diffmap-ccn-genes, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center"}
cell_cycle_genes <- !is.na(fData(scData)$cell_cycle_phase)
sum(cell_cycle_genes)
most_exprs_ccycle <- which(cell_cycle_genes & 
                               fData(scData)$total_feature_counts > 100000)
p1 <- plotPCA(scData, feature_set = cell_cycle_genes, 
               colour_by = featureNames(scData)[most_exprs_ccycle[8]], 
               shape_by = "sample_type", 
               size_by = "use" ) + ggtitle("PCA")
p2 <- plotDiffusionMap( scData, feature_set = cell_cycle_genes, 
                        colour_by = featureNames(scData)[most_exprs_ccycle[8]], 
                        shape_by = "sample_type", 
                        size_by = "use" ) + ggtitle("Diffusion map")
multiplot(p1, p2, cols = 2)

@@

One element of QC that can prove challenging to judge from the above criteria is if more than one cell ( or cell debris from other cells ) is being sequenced. With most cell processing platforms a small percentage of cells won't be truly single. One solution is imaging individual cells,  which also allows for the incorporation of fluorescence phenotypes such as viability and nuclear content ( for cell cycle analysis ). Using the [EBImage]( http://bioconductor.org/packages/release/bioc/html/EBImage.html ) package,  an example image of a stained cell is processed and displayed below,  with results incorporated into the sample meta-data.  
\

@@

## 4. QC THE GENES

@@

It can also be useful to look at total expression levels across gene biotypes. 
The plot below shows the distributions of log-10 scale total gene counts for 
genes of different annotated biotypes. The biotypes are ordered by median total
counts, and the jittered points beneath the violin plots indiciate the number of 
genes in the biotype. We want to see many protein-coding genes with reasonably 
high expression. This is exactly what we see here, with protein-coding genes 
appearing as the biotype with second-highest median expression behind the very 
small (but highly expressed) category of Mt_rRNA genes.
\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=8, fig.height=6, fig.align="center"}
oo <- order(unlist(dplyr::summarise(dplyr::group_by(fData(scData), gene_biotype),
                 median(log10_total_feature_counts))[,2]), decreasing = TRUE)
fData(scData)$biotype <- factor(
    fData(scData)$gene_biotype, 
    levels = levels(as.factor(fData(scData)$gene_biotype))[oo])

@@

```{r append-gene-symbols}
fData(scData)$gene_symbol <- fData(scData)$hgnc_symbol
fData(scData)$gene_symbol[is.na(fData(scData)$gene_symbol)] <-
    featureNames(scData)[is.na(fData(scData)$gene_symbol)]
featureNames(scData) <- paste(fData(scData)$hgnc_symbol, featureNames(scData),
                              sep = "_")
```

```{r, echo=T, eval=T, message=F, warning=F, fig.width=6, fig.height=6, fig.align="center"}
plotQC( scData[fData( scData )$use, pData( scData )$use], 
        type = "highest-expression", 
        col_by_variable = "sample_type", feature_names_to_plot = "gene_symbol")
```

@@

## 5. NORMALISE THE DATA

@@

Experimental design is a critical, but neglected, aspect of \gls{scrnaseq} studies. To the best of our knowledge, methods like those described in this section for exploring experimental and QC variables and the experimental design, do not feature in any scRNA-seq software packages apart from `r Biocpkg("scater")`. There are a very large number of potential confounders, artifacts and biases in \gls{scrnaseq} studies. Exploring the effects of such explanatory variables (both those recorded during the experiment and computed QC metrics) is crucial for appropriate modeling of the data. The \scater\ package provides a set of methods specifically for quality control of experimental and explanatory variables, which will be demonstrated briefly here.

### Finding important explanatory variables

@@

### Scaling normalisation

In `r Biocpkg("scater")` it is easy to perform scaling normalisation using the "TMM" approach (the default in the [edgeR]() package), the "RLE" approach (default in the [DESeq]() package), an "upper-quartile approach" (as proposed by Bullard et al (2010)) or simple scaling by total counts. Size-factor normalisation can be carried out on a subsetted SCESet object, normalising either to ERCC spike-in genes or to endogenous genes if desired. Under certain circumstances either may be appropriate. Careful thought should be given to applying scale-factor normalisation as underlying assumptions from methods developed for bulk RNA-seq may not be appropriate for all single-cell datasets.

However, [Lun et al (2016)](http://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0947-7) recently published a size-factor normalisation method specifically designed
for scRNA-seq data. This method performs much better on single-cell data and its
implementation in the Bioconductor package `r Biocpkg("scran")` allows seamless integration 
into the `r Biocpkg("scater")` workflow. (The `r Biocpkg("scran")` package itself depends on `r Biocpkg("scater")`).

We recommend using the `r Biocpkg("scran")` size-factor normalisation approach and 
demonstrate its use here.

@@

isSpike(scDataQC) <- "spikes"
scDataQC_spikenorm <- computeSpikeFactors(scDataQC)
summary(scDataQC_spikenorm$size_factor)
scDataQC_spikenorm <- normalise(scDataQC_spikenorm)

@@

We would prefer such technical effects to be removed by normalisation, so below we demonstrate a further alternative approach. We see how "customised" normalisation approaches can be easily incorporated into the `r Biocpkg("scater")` workflow.

### Expression-conditional normalisation

In this example we normalise (by cell) and standardise (by gene) counts conditioned on expression level, showing the PCA for the corrected data. One advantage of this approach is that a biologically 'noisy' gene is naturally defined as one with greater dispersion than other genes at a similar expression level. In the normalised data these are genes with a mean absolute deviation greater than 1.

@@

### Regressing out known sources of variation

@@

```{r regress-factors-norm,  echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=6}
design <- model.matrix(~scDataQC$log10_counts_endogenous_features +
                           scDataQC$pct_counts_top_100_features +
                           scDataQC$pct_counts_feature_controls +
                           scDataQC$c1_machine)
set_exprs(scDataQC, "norm_exprs_resid") <- norm_exprs(
    normaliseExprs(scDataQC, design = design,
                   method = "none", exprs_values = "exprs",
                   return_norm_as_exprs = FALSE) )

@@

Here, regressing out only the technical QC metrics does not remove the major 
effect of C1 machine, but regressing out cDNA concentration as well does (so 
the first principal component now separates the two patients). Thus, for 
removing known effects, the regression approach implemented in `r Biocpkg("scater")` is 
effective, but more needs to be done to find _hidden_ effects such as unknown
batch effects, or similar.

### Normalisation with `removeBatchEffect` from `limma`

@@

### Batch correction with `RUVSeq`

@@

scDataRUV <- filter(scData, use)
scDataRUV <- scDataRUV[fData(scData)$use | fData(scData)$is_feature_control,]
scIdx <- matrix(-1, ncol = max(table(scDataRUV$sample_type)), nrow = 2)
scIdx[1,] <- which(scDataRUV$sample_type == "cell from patient B")
tmp <- which(scDataRUV$sample_type == "cell from patient A")
scIdx[2, 1:length(tmp)] <- tmp
# cIdx <- (fData(scDataRUV)$is_feature_control & fData(scDataRUV)$n_cells_exprs >= 10) 
## Running RUVs
cIdx <- rownames(scDataRUV)
ruv_list <- list()
for (k in 1:4) {
    ruvs <- RUVs(counts(scDataRUV), cIdx, k = k,
                 scIdx = scIdx, isLog = FALSE)
    norm_counts(scDataRUV) <- ruvs$normalizedCounts
    exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                           prior.count = scDataRUV@logExprsOffset)
    plt_norm_ruv <- plotPCA(
        scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = TRUE) + 
        ggtitle(paste("PCA - RUVs normalisation: k =", k)) + 
        theme_bw(8) + theme(legend.position = "bottom")
    ruv_list[[k]] <- plt_norm_ruv
}

figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list, labels = letters, ncol = 2)
figplot_ruv
cowplot::save_plot("../figures/supplement/ruvs_figure.png", figplot_ruv, 
                   ncol = 3, nrow = 3, base_height = 5, base_aspect_ratio = 0.9)
```

Just removing the first latent variable found with RUVs now effectively splits cells from the two patients in the first principal component of the adjusted expression values.

The QC plot below confirms that the first principal component is now highly correlated with patient.

```{r ruv-seq-s-qc-plots, echo=TRUE, eval=TRUE, message=FALSE, results='hide'}
plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
```

@@

## 6. DOWNSTREAM ANALYSIS

@@

In `r Biocpkg("scater")` the `plotExpression` function enables the convenient visualisation of expression values for a set of features. Here, the expression values for the six most DE genes for expression frequency between patients are shown. The units for expression in the plot can be defined with the `exprs_values` argument (the expression values must exist with the provided name in the `assayData` slot of the SCESet object; if not the default `exprs` values will be used, with a warning). As with other plots in `r Biocpkg("scater")` we can use phenotype data variables to define the colour and shapr of the points.

@@

## 7. SESSION INFORMATION

@@

```{r figures-for-paper, echo=TRUE, results='hide'}
## Figures for the paper
if ( requireNamespace("cowplot") ) {
    ## QC figure
    scData$well_type <- "single cell"
    scData$well_type[scData$sample_type == "blank (empty site) control"] <-
        "blank control"
    scData$well_type[scData$sample_type == "bulk control from patient A"] <-
        "bulk control"
    scData$well_type[scData$sample_type == "bulk control from patient B"] <-
        "bulk control"
    p1 <- plot( scData, colour_by = "well_type", exprs_values = "counts") +
        theme(legend.position = c(0.8, 0.25))
    p2 <- plotPCA( scData, shape_by = "sample_type", 
                   pca_data_input = "pdata", detect_outliers = TRUE,
                   selected_variables = c("cdna_recovered_in_ng_per_ul_no_miss",
                                          "pct_counts_top_100_features",
                                          "total_features", 
                                          "pct_counts_feature_controls",
                                          "n_detected_feature_controls", 
                                          "log10_counts_endogenous_features",
                                          "log10_counts_feature_controls"))
    p3 <- plotQC( scData[fData( scData )$use, pData( scData )$use], 
        type = "highest-expression", feature_names_to_plot = "gene_symbol",
        col_by_variable = "sample_type", n = 20) + coord_cartesian(xlim = c(0, 6)) +
        theme(axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 11))
    p4 <- plotQC(scData, type = "exprs")
    p5 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "explanatory-variables", 
        variables = c("pct_counts_top_100_features", "total_features", 
                      "c1_machine",
                      "n_detected_feature_controls", 
                      "log10_counts_endogenous_features",
                      "sample_type") ) + 
        xlab(expression(paste(R ^ 2, " (% variance explained; log10-scale)")))
    p6 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "c1_machine")
    figplot1 <- cowplot::plot_grid(p1, p2, p3, p4, p5, p6, 
                                  labels = letters, ncol = 3,
                                  label_size = 20)
    cowplot::save_plot("../figures/figure3.pdf", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3.png", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3.jpg", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4, dpi = 350)

    ## alternative version for better readability
    figplot1_alt <- cowplot::ggdraw() +
        cowplot::draw_plot(p1, 0.01, .5, 0.28, .48) +
        cowplot::draw_plot_label(letters[1], x = 0, y = 1, size = 20) +
        cowplot::draw_plot(p2, 0.29, .5, 0.35, .48) +
        cowplot::draw_plot_label(letters[2], x = 0.29, y = 1, size = 20) +
        cowplot::draw_plot(p4, 0.64, .5, 0.36, .48) +
        cowplot::draw_plot_label(letters[3], x = 0.63, y = 1, size = 20) +
        cowplot::draw_plot(p3, 0.01, 0, 0.5, .48) +
        cowplot::draw_plot_label(letters[4], x = 0, y = 0.49, size = 20) +
        cowplot::draw_plot(p5, 0.52, 0, .48, .48) +
        cowplot::draw_plot_label(letters[5], x = 0.515, y = 0.49, size = 20)

    cowplot::save_plot("../figures/figure3_alt.pdf", figplot1_alt, ncol = 3, nrow = 2,
                       base_height = 3.2, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3_alt.png", figplot1_alt, ncol = 3, nrow = 2,
                       base_height = 3.2, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3_alt.jpg", figplot1_alt, ncol = 3, nrow = 2,
                       base_height = 3.2, base_aspect_ratio = 1.4, dpi = 350)

    ## save find-pcs plot to add to supplement
    cowplot::save_plot("../figures/find-pcs_c1_machine.pdf", p6,
                       base_height = 5, base_aspect_ratio = 1.1)
    cowplot::save_plot("../figures/find-pcs_c1_machine.png", p6,
                       base_height = 5, base_aspect_ratio = 1.1, dpi = 350)
    cowplot::save_plot("../figures/find-pcs_c1_machine.jpg", p6,
                       base_height = 5, base_aspect_ratio = 1.1, dpi = 350)

    ## dimension reduction figure
    ww <- which(featureNames(scDataQC) %in% rownames(de)[c(1:2, 5)])
    # ww <- which(featureNames(scDataQC) %in% c("IGHM_ENSG00000211899", 
    #                                           "IGHV4-34_ENSG00000211956",
    #                                           "CD86_ENSG00000114013"))
    # ww <- which(featureNames(scDataQC) %in% c("IGHM", 
    #                                           "IGHV4-34",
    #                                           "CD86"))
    featureNames(scDataQC)[ww] <- as.character(fData(scDataQC)$hgnc_symbol[ww])
    feats_to_plot <- as.character(fData(scDataQC)$hgnc_symbol[ww])
    featureNames(scData)[most_exprs_ccycle[7]] <- 
        fData(scData)$hgnc_symbol[most_exprs_ccycle[7]]
